---
title: "Factor Lab"
author: "Jacqui Buchanan"
date: "February 25, 2016"
output: html_document
---

### FACTORS.


##### Factors are one of the four primary "classes" in R, and are categories that represent groups within your data.

##### It may seem like factors are redundant when you can numerically create levels (for example instead of casting "treatment" and "control" as factors, we could assign 1=treatment and 0=control), factors are useful because they save memory and increase processing speed. 

##### Factors also take advantage of R's object-oriented'ness, and lets you perform analysis by groups more easily.

##### Important note: Factors canNOT be directly re-cast from numeric. You have to cast them as a character, and then as a number.

```{r}
x <- as.factor(c(9,9,9,11,10,12,12,12,10,10,11,12))
x
class(x)
as.numeric(x)
x
```

##### Oh no! We've totally lost the original x.

```{r}
x <- as.factor(c(9,9,9,11,10,12,12,12,10,10,11,12))
x
x2 <- as.numeric(as.character(x))
x2
```

##### Sometimes (as we all know from the first semester), R is sneaky and through a function will turn one data class into another. 

##### An example of this is the cut function, c( ), which is used to convert a numeric variable into a factor. 

##### The breaks= argument to cut is used to describe how ranges of numbers will be converted to factor values.

##### If a number is provided through the breaks= argument, the resulting factor will be created by dividing the range of the variable into that number of equal length intervals; if a vector of values is provided, the values in the vector are used to determine the breakpoint.

##### Here's an example: 

```{r}
y <- 1:12
new.y <- cut( y, breaks=3, labels=NULL )
new.y
class(new.y)
```

##### As you can see, new.y has been changed to a factor because it was divided into groups with the cut function. This is problematic if we want to perform any mathematical functions using new.y.

```{r}
this.ex <- new.y*3
this.ex
```

##### So take care if you're using cut or other functions that group, and make sure you either want your data in factor form, or you re-cast it at some point.

```{r}
data("warpbreaks")

levels( warpbreaks$tension )

summary( warpbreaks$tension )

```

##### R is grouping the wool tension by these levels (low, medium, high).  Internally, R stores the integer values like this: 1=L, 2=M, and 3=H.  

##### If we do a summary of this variable, R tallies each category (see above). For factors like this, R won’t let us do a mean or any other statistic of a factor variable other than a count. 


### I'm sure there are many more frustrating (read:useful) ways to use factors, this was just the tip of the iceberg. Enjoy!































